Documenting Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models

### 2022-11-23 manuscript/manuscript.rmd
Working on the manuscript

  \noindent Abbreviations used: United States (US), Coronavirus Disease 2019 (COVID-19), Generalized
  Additive Model (GAM), Centers for Disease Control and Prevention (CDC)
bibliography: bibliography.bib
csl: apa.csl
link-citations: true
urlcolor: blue
linkcolor: black
---

@@

In the US context, one of the key aspects to the geographic story of COVID-19's
spread and diffusion was the early surge of cases and epicenter in New York City
during March 2020 [@thompson_covid-19_2020] followed by subsequent waves of cases in the
South and Midwest [@shumaker_covid-19_2020; @scott_these_2020; @glenza_covid_2020]. As Park et al. 
stated in their article *COVID-19 Deaths in the United States: Shifts in Hot Spots over the Three Phases of the Pandemic and the Spatiotemporally Varying Impact of Pandemic Vulnerability* where they summarized
US trends from March 2020 to May 2021, "hot spots have shifted from densely populated cities and the states with a high percentage of socially vulnerable individuals to the states with relatively relaxed social distancing requirements, and then to the states with low vaccination rates" [@park_covid-19_2021].

@@

<!-- 
Want to cite: 
* Mude et al (racial disparities meta-analysis);
* Chen & Krieger on their article "Revealing the Unequal Burden" 
* Hanage et al COVID-19: US federal accountability for entry, spread and inequities
* Kim et al on COVID-19 magnifying the effect of health disparities (esp. re: testing (re: care providers, transportation equity))
* Romano et al (CDC MMWR on trends and disparities in hospitalizations by race/ethnicity and region)
* Okonkwo et al: accelerating health inequities
* Del Rios: COVID-19 is an inverse equity story, not a racial equity success story
* Goldstein/NYT: During the Omicron surge, Black New Yorkers were hospitalized at a rate more than twice that of white residents.
* Mueller/NYT In Rural America, Covid Hits Black and Hispanic People Hardest
* Hill et al / KFF COVID-19 Cases and Deaths by Race/Ethnicity: Current Data and Changes Over Time
* Lo et al The polarization of politics and public opinion and their effects on racial inequality in COVID mortality
* Johnson and Keating: Whites now more likely to die from covid than Blacks: Why the pandemic shifted 
* Kaashoek The Evolving Roles of US Political Partisanship and Social Vulnerability in the COVID-19 Pandemic from February 2020 - February 2021
* 
* Krieger et al on a warning against static measures

Outline:

Disparities became immediately obvious: 
  Though many have repeated the oft-quoted line "viruses do not discriminate,"
  it was immediately clear that COVID-19 would lay bare the sharp inequities
  that exist in tge US; almost immediately some groups were shown to be
  experiencing higher rates of COVID-19 infection and mortality. Those groups
  have been shown to include marginalized racial/ethnic groups, those who do not
  speak English, transportation insecurity, the elderly, those exposed to
  chronic pollution, those living in crowded housing or areas with financial
  insecurity, and those with disabilities, housing instability and/or
  homelessness, pre-existing conditions (including chronic obstructive pulmonary
  disease, diabetes, heart diseases, etc.).  Smoking.  Obesity.  Acute Kidney
  Injury.

There are some aspects of this story that we still don't know: 
  The details on COVID-19 rates are not availability stratified by both disability
  and race/ethnicity simultaneously making it difficult to understand how much of
  racial/ethnic disparities in COVID-19 are attributable to increased disability
  rates among historically (and even now) marginalized racial/ethnic groups.

  Moreover, it is known that many of the risk factors for increased COVID-19 severity 
  are prevalent at increased rates among sexual minority persons compared to heterosexual
  persons, suggesting that sexual minority persons may be at increased risk of 
  severe COVID-19.

Vaccines didn't right the boat: 
  Again, when vaccines became available, I think many thought "surely this will
  be the silver bullet that solves the problem";  inequities in who were able to
  get vaccine appointments reinforced existing patterns in internet access,
  technological literacy and availability of transportation. Even after the
  initial vaccine rollout, inequities persist in vaccination clinic locations
  with 'vaccine deserts' being more likely to be located in less affluent, less 
  white communities.

Throughout this timeline, it has become increasingly apparent that local-level
politics represents a critical factor influencing the trajectory of COVID-19.

COVID is not a racial equity success story: 
  Finally, racial inequities in COVID-19 in the US have been widening and
  shrinking during the course of the pandemic, largely widening during surges
  and shrinking when case rates and mortality have been lower.  During Fall 
  2022, the age-adjusted COVID-19 mortality rate for white non-Hispanic 
  people surpassed the mortality rate for Black or African/American people. 
  This has been explained as being due to those who still remain unvaccinated
  being more likely to be Republican and white. While reducing racial inequities
  in COVID-19 mortality rates during Fall 2022, may sound like a good thing, the
  cumulative burden of COVID-19 leaving in its wake long-COVID, lost loved ones,
  jobs lost, and damage to individuals' health that may never be reversed that
  disproportionately falls on marginalized racial/ethnic groups does not
  represent a health equity success story.
-->

  - Insert section on prior modeling results
  - Insert discussion motivating the need for spatiotemporal modeling

@@

  * Counts of COVID-19 deaths [@the_new_york_times_coronavirus_2021].
  * Population size estimates for 2020 from the U.S. Census (Redistricting 
  File Public-Law 94-171 Dataset) [@us_census_bureau_2020_2021].
  * Median age, median household income, racial/ethnic composition, population
  density, percent below the federal poverty line, and number of households with
  high (\$100k+)/low (<\$25k) household income by racial/ethnic group from the
  2014-2019 5-year American Community Survey [@us_census_bureau_american_2020]
  through the `tidyverse` R package [@walker_tidycensus_2022]. 
  * Votes cast in the 2020 presidential election [@mit_county_2022]

@@

Generalized additive models (GAMs) improve upon generalized linear models by
allowing for the fitting of smooth functions that transform the $x$ variables.
This is a convenient means to include nonlinear relationships between the
outcome and predictor variables. Whereas a generalized linear model may have looked like 
$$ g(\mu_i) = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} ...$$ 
a generalized additive model could look like 
$$ g(\mu_i) = \mathbf A_i \mathbf \theta + f_1(x_{i1}) + f_2(x_{i2}) + f_3(x_{i3}, x_{i4}) + \dots$$
where $\mu_i \equiv \mathbb E(Y_i)$, $\mathbf A_i$ is a row of the model matrix 
for any strictly non-parametric model components, $\mathbf \theta$ is the corresponding
parameter vector, and the $f_j$ are smooth functions of the covariates $x_k$ [@wood_generalized_2017].

@@

@@

  * Missing data 
  * Ecological fallacy 
  * Age effects
  * Fixed population denominators -> real-time denominator estimates
  * Vaccination and mobility effects

>>>>

  \noindent Abbreviations used: United States (US), Coronavirus Disease 2019 (COVID-19), Generalized
  Additive Model (GAM), Centers for Disease Control and Prevention (CDC)
bibliography: bibliography.bib
csl: apa.csl
link-citations: true
urlcolor: blue
linkcolor: black
header-includes: 
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
---

@@

In the US context, one of the key aspects to the geographic story of COVID-19's
spread and diffusion was the early surge of cases and epicenter in New York City
during March 2020 [@thompson_covid-19_2020] followed by subsequent waves of cases in the
South and Midwest [@shumaker_covid-19_2020; @scott_these_2020; @glenza_covid_2020]. As Park et al. 
stated summarized trends in the US from March 2020 to May 2021, "hot spots have shifted from densely populated cities and the states with a high percentage of socially vulnerable individuals to the states with relatively relaxed social distancing requirements, and then to the states with low vaccination rates" [-@park_covid-19_2021].

@@

Even since the beginning of the COVID-19 outbreak in the US, data reflected
sharp inequities in mortality rates.  During January 22nd to May 5th 2020,
county COVID-19 mortality rates were 4.94 (95% CI 4.78, 5.09) times higher in
counties in the highest quintile of percent People of Color (61%-100%) compared
to counties in the lowest quintile of percent People of Color (0%-17.2%)
[@chen_revealing_2021].  This was not wholly unanticipated: as COVID-19 was
beginning to take off in the US, some were already calling attention to the fact
that COVID-19 threatened to exacerbate existing disparities
[@kim_covid-19_2020]. Kim, Marrast, and Conigliaro noted three structural
barriers in COVID-19 prevention and care: 1) originally requiring New York
residents to have a doctor's prescription for a COVID-19 test reduced the
opportunity for healthcare for People of Color as they are less likely to have a
primary care provider; 2) drive-through testing made testing disadvantaged those
who rely on public transportation; 3) quarantining at home while waiting the
7-10 days originally required for COVID-19 test results to come back posed an
economic and social challenge that many in already financially difficult
situations may not have been able to take on [-@kim_covid-19_2020]. Others noted
yet more reasons why COVID-19 threatened to worsen an already inequitable
healthcare landscape in the US: those who reside in prisons and jails,
immigrants and undocumented people, people with disabilities, and people
experiencing homeless all face additional challenges in seeking and getting the
healthcare they deserve [@okonkwo_covid-19_2021]. It's not just a hypothetical 
that those with more structurally enfranchised privileges have been more able to
mitigate their risk of negative health outcomes associated with COVID-19 
— during February 1 to April 1, 2020, New York City residents who lived in more affluent 
neighborhoods were more likely to have left the city, while New York City residents
from more Black and Hispanic neighborhoods were more likely to continue working [@coven_disparities_2020].

> Contrary to the oft used phrase that the ‘virus does not discriminate’, the
data presented here suggest that this virus, as many other infectious diseases,
has the greatest implications for the most vulnerable people. The intersections
between health and human rights are clear—the health of a society and
vulnerability to a pandemic are directly related to its human rights track
record for those who are marginalised. [@okonkwo_covid-19_2021]

When vaccines became available, vaccine appointments were often only available
to be scheduled through online web-portals contributing to the inequities
between those who had internet access and technological literacy and those who
didn't [@press_inequities_2021]. Vaccination sites have not been equally
distributed and areas determined to be vaccine deserts have been found to have
disproportionately more Black and Hispanic residents [@rader_spatial_2021]. In
fact, healthcare facilities in counties with higher Black composition had 32%
(95% CI 14%-47%) lower odds of serving as vaccine sites
[@hernandez_disparities_2022].

What vaccination has been administered hasn't suddenly erased the unequal burden
of COVID-19 either; in August 2022 the New York Times was reporting "Black death
rates at this winter’s peak were greater than those of white people by 34
percent in rural areas, 40 percent in small or medium cities and 57 percent in
big cities and their suburbs" [@goldstein_during_2022]. As COVID-19 case and
mortality rates have waxed and waned, the inequities have widen and shrunken,
often with racial/ethnic inequities growing during times when COVID-19 rates
have surged [@hill_covid-19_2022]. During 2022, the age-standardized
COVID-19 mortality rates for white people have, at times, been slightly 
higher than those of Black and Hispanic people, predominantly because the 
mortality rate among white people has increased [@johnson_whites_2022]. 
It's important to note that white COVID-19 mortality rates overtaking 
the Black mortality rates does not imply that equity has been established: neither
does this undo the cumulative impact of mortality (which has been twice as high for
Black people compared to white people [@hill_covid-19_2022]), nor does it imply 
the underlying systemic barriers to equity have been overturned
[@del_rios_covid-19_2022]. As Del Rios, Chomilo, and Lewis note, instead,
COVID-19 leaves in its wake more years of life expectancy lost, wages lost, and
degradation of mental health in Communities of Color.

@@

  * Counts of COVID-19 deaths [@the_new_york_times_coronavirus_2021].
  * Population size estimates for 2020 from the U.S. Census (Redistricting 
  File Public-Law 94-171 Dataset) [@us_census_bureau_2020_2021].
  * Median age, median household income, racial/ethnic composition, population
  density, percent below the federal poverty line, and number of households with
  high (\$100k+)/low (<\$25k) household income by racial/ethnic group from the
  2014-2019 5-year American Community Survey [@us_census_bureau_american_2020]
  through the `tidyverse` R package [@walker_tidycensus_2022]. 
  * Votes cast in the 2020 presidential election [@mit_election_data_and_science_lab_county_2022]

@@

In contrast, generalized additive models (GAMs) improve upon generalized linear models by
allowing for the fitting of smooth functions that transform the $x$ variables.
This is a convenient means to include nonlinear relationships between the
outcome and predictor variables. Whereas a generalized linear model may have looked like 
$$ g(\mu_i) = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} ...$$ 
a generalized additive model could look like 
$$ g(\mu_i) = \mathbf A_i \mathbf \theta + f_1(x_{i1}) + f_2(x_{i2}) + f_3(x_{i3}, x_{i4}) + \dots$$
where $\mu_i \equiv \mathbb E(Y_i)$, $\mathbf A_i$ is a row of the model matrix 
for any strictly non-parametric model components, $\mathbf \theta$ is the corresponding
parameter vector, and the $f_j$ are smooth functions of the covariates $x_k$ [@wood_generalized_2017].

```{r gam-demo-figure, fig.cap = "A demonstration of fitting regression models using B-spline basis functions. Part A. shows the simulated dataset. Part B. shows the B-spline basis functions. Part C. shows the weighting of the B-spline basis functions and their weighted sum with an intercept added to create the regression model fit to the data."}

knitr::include_graphics(here("analysis/10_supplementary_gam_figures/gam_demo_figure.png"))
```

The smooth functions estimated as part of fitting a GAM are constructed using
spline basis functions. These spline basis functions allow for the smooth
interpolation of trends in the data allowing for the incorporation of
nonlinearity. To avoid overfitting the data, a penalty term is introduced that
controls the degree of "wiggliness" or smoothness and this penalty term is fit
using generalized cross validation. B-splines are one kind of spline basis function 
that is commonly used and are especially popular because of their property that they 
are non-negative on only a finite interval.

While a variety of varying kinds of spline bases exist (cubic splines, B- and
P-splines, thin-plate splines, etc.), our particular application setting
warrants the use of tensor-product smooths because we require anisotropic
penalization. That is to say, when we seek to create spatiotemporally smoothed
model estimates, it's inappropriate to assume that the amount of smoothing
across space should be the same as the amount of smoothing across time because
they're in fundamentally different units (space being measured in Cartesian
coordinates and time being measured in years, months, days, etc.). More details
about tensor-product smooths are available in Simon Wood's *Generalized Additive Models - An Introduction with R* [-@wood_generalized_2017].

## Overdispersion and Event Modeling

Above and beyond using the GAM framework to allow for flexible, nonlinear
relationships between our observed county-level variables and COVID-19
mortality, we must have a model specification that agrees with the data
generating process. In our case, the data generated are counts of deaths per
population, and the class of models most suited to represent counts of events
are Poisson, quasi-Poisson, and negative binomial models. Here we've chosen to 
use the negative binomial model as it accounts for overdispersion and has 
the intuitive interpretation of a Poisson model with gamma distributed underlying
rate parameter [@gelman_bayesian_2013].

A negative binomial model in the context of modeling count data can be written
$$y_i \sim \text{NegativeBinomial}(u_i \theta_i, \phi),$$
where $y_i$ are the counts observed, $u_i$ is the exposure, $\theta_i$ are
the rates, and $\phi$ determines the amount of overdispersion. The rates
are modeled as $\theta_i = e^{X_i \beta}$ where $X_i$ are the observed
covariates and $\beta$ are the coefficients on the covariates corresponding
to log rate ratios. The logarithm of the exposure, $\log(u_i)$ is often called 
(and later herein referenced as) the offset. In epidemiological contexts, the offset 
is often representative of the amount of person-time during which observations were recorded. Whereas the Poisson model holds that $\text{var}(y) = \mu$ where $\mu$
is the average rate, the negative binomial model instead assumes that
$\text{var}(y) = \mu + \mu^2/\phi$. The Poisson model
is a special case of the negative binomial model when $\phi \to \infty$.

An alternative and equivalent formulation of the negative binomial that is commonly 
used makes the connection to the Poisson model even more clear: instead of 
$\theta_i$ and $\phi$, using $\alpha$ and $\beta$, 
$$y \sim \text{NegativeBinomial}(\alpha, \beta), \text{ and}$$
$$\text{NegativeBinomial}(y|\alpha, \beta) = \int \text{Poisson}(y|\theta) \text{Gamma}(\theta | \alpha, \beta) d\theta.$$

Introductions to and additional exposition on the negative binomial model,
especially in the context of modeling counts of events outcome data, are
available in *Bayesian Data Analysis* and *Regression and Other Stories*
[@gelman_regression_2020; @gelman_bayesian_2013].

## Variables of Interest

Besides age, which is the greatest risk factor for COVID-19 mortality, we have
included the following variables as covariates of interest:

  * Population Density per Square Mile
  * Median Household Income
  * Proportion in Poverty
  * the Index of Concentration at the Extremes for Racialized Economic Segregation 
  * Political Lean in the 2020 Election (1 = 100% Republican votes, -1 = 100% Democratic Votes)

#### Index of Concentration at the Extremes for Racialized Economic Segregation

The Index of Concentration at the Extremes (ICE) is a measure which describes
how concentrated a given area's population is in terms of the extreme ends of
privilege and marginalization [@krieger_public_2016]. 
In general, the ICE measure is formulated as

$$ \text{ICE} = \frac{\text{Number of People in Most Privileged Category} - \text{Number of People in Least Privileged Category}}{\text{Total Population}}$$

Applying the ICE approach to a specific context involves defining the axes of
privilege of interest.  In this case, data on racialized economic segregation
are used from the US Census American Community Survey on White households
earning more than $100,000 a year (the most privileged group) or households of
People of Color earning less than $25,000 a year (the least privileged group).
This variable is referred to throughout as the ICE for racialized economic
segregation, or `ICEraceinc` in the code.

Compared with the Gini
coefficient which is one of the most popular methods for summarizing area-level
rates of inequities, the ICE the major advantage that it is more suitable for
describing inequities at smaller area levels [@krieger_public_2016]. While the
Gini coefficient measures within-area dissimilarity (as in, for example, how
unequal wealth is distributed within a county), the ICE measure establishes
where on a spectrum a given county's population falls allowing for comparison
across counties. The Gini coefficient suffers from the fact that areas which are
made up of relatively homogeneous populations will appear as having low
within-area inequality (and therefore low Gini coefficient). Instead, the ICE
measures how much of the population is either privileged or not. The Gini
coefficient remains useful for reporting on the degree of inequity in larger
areal units (like countries, states, and regions), but at smaller areal units
(like counties, ZIP codes, census tracts) can be more difficult to interpret and
compare. Maps of ICE measure can elucidate what spatial social segregation and
polarization exist, and the ICE for racialized economic segregation has been
repeatedly and significantly associated with COVID-19 outcomes
[@hanage_covid-19_2020; @chen_revealing_2021; @krieger_relationship_2022;
@eichenbaum_health_2022; @brown_ecological_2021; @sonderlund_racialized_2022;
@saha_neighborhood-level_2020].

#### Political Lean

Political lean has been associated with COVID-19 mortality in numerous studies,
with plausible mechanisms explaining the association including differences in
non-pharmaceutical intervention uptake (mask usage, social distancing,
quarantining), differences in rates of vaccination, differences in political
leadership's messaging, resource allocation, and the adoption of policy
interventions [@krieger_relationship_2022; @kaashoek_evolving_2021; @leonhardt_red_2021; @gonzalez_conservatism_2021; @grossman_political_2020].

@@

## Application 1: Non-Spatial Covariate Effects Over Time

The GAM models are fit with the following structure using 
the `gam` function from the `mgcv` package [@wood_mgcv_2022]:

```{r gam-demo-code, echo = TRUE, eval=FALSE}
gam(
  formula = deaths ~ s(median_age) + te(covariate, date, d=c(1,1)), # regression formula
  offset = log(popsize/1e5/12), # our offset represents the person-time
  data = df,     # our data-set of county-level observations
  family = nb(link='log')  # indicates negative binomial family and a log-link function
)
```

The formula used puts a one-dimensional smoothing spline to represent a
nonlinear age-effect and a two-dimensional tensor-product smooth on the
interaction between the given covariate and the date. The `d=c(1,1)` argument
provides the instruction necessary to consider covariate and date as being on
separate scales and therefore to fit the tensor-product smooth with anisotropy —
that is, to allow for independent amounts of scaling in the dimensions of the
covariate and time. The offset used structures the regression to estimate rates
in units of person-time per 100,000 person-years. Since the death counts are
aggregated to the monthly level, the person-time in units of 100,000
person-years are calculated by taking each county's population size, dividing by
100,000, and dividing by 12 (for the 12 months in a year).

The above structure is used to estimate models for our different `covariate`
variables of interest: median income, percent in poverty, the ICE for racialized
economic segregation, political lean.  Whereas the other models treat median age
as a variable to adjust for, the model presenting the effect of median age over
time treats median age as the main covariate of interest including it as the 
`te(covariate, date, d=c(1,1))` and dropping the `s(median_age)` term which is 
redundant. Results of these models are summarized in \@ref(fig:one-variable-over-time)
where the additional COVID-19 mortality associated with each covariate is visualized.

(ref:medianage) [@us_census_bureau_nation_2022]

```{r one-variable-over-time, fig.cap = "Additional COVID-19 mortality associated with covariates over time. A) age and time, B) log (base 10) population density and time, C) median income and time, D) proportion in poverty and time, E) ICEraceinc and time, F) political lean and time. In panels B-F the effect of age is marginalized out using the median age in the US, 38.8 (ref:medianage).", fig.align = 'center', out.height = '75%'}
knitr::include_graphics(here("analysis/04_one_variable_over_time/panel_figure.png"))
```

Likelihood ratio tests confirmed that models with covariates interacted with
time had significantly lower residual deviance ($p \leq 2.2e^{-16}$ for all
models) compared to models only including a spline term on median age and the
given covariate not interacted with time. The model interacting median age 
and time was compared to a model with a spline for median age alone. Akaike Information
Criteria values were also lower for all models compared to models that did
not interact the covariates with time.

Another non-spatial application of GAMs to describe the distribution of 
COVID-19 mortality in the US over time is to consider the associations of 
mortality with three-way interactions of time and two covariates taken together. 
In the following example, the formula used is `deaths ~ te(date, ICEraceinc, median_age, d=c(1,1,1))`.  Again, the `d` argument specifying the marginal 
basis dimensions is used indicate that each of the `date`, `ICEraceinc`, and 
`median_age` measures are in different units and should not be smoothed assuming
that a one unit difference in one variable is comparable to a one unit difference
in another variable.

```{r two-variables-over-time, fig.cap="The changing interacted effect of the ICE for racialized economic segregation and age over time. A) May 2020, B) November 2020, C) May 2021, D) October 2021, E) February 2022, F) August 2022. An animated version is available online at "}
knitr::include_graphics(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/panel_figure.png"))
```

@@

  * Ecological fallacy 
  * Age effects
  * Fixed population denominators -> real-time denominator estimates
  * Vaccination and mobility effects
  * Equal Smoothing Across All Areas
  * Soap Film Smoothers
